---
title: 'Analysis of the Fibrosis datataset'
author: "Hector Roux de Bézieux"
bibliography: ref.bib
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Analysis of the Fibrosis dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r packages, include=F}
library(knitr)
opts_chunk$set(
  fig.pos = "!h", out.extra = "", warning = F, 
  fig.align = "center"
)
```

# Load data

The dataset we will be working with concerns a single-cell RNA-sequencing dataset consisting of three cancer models under a  KRAS(G12C) inhibition [@Xue2020].

```{r}
libs <- c("here", "tidyr", "dplyr", "Seurat", "scater", "slingshot", "cowplot",
          "condiments", "ggplot2", "readr", "tradeSeq",  "pheatmap", "fgsea",
          "msigdbr", "scales")
suppressMessages(
  suppressWarnings(sapply(libs, require, character.only = TRUE))
)
rm(libs)
theme_set(theme_classic() +
            theme(rect = element_blank(),
                  plot.background = element_blank(),
                  strip.background = element_rect(fill = "transparent")
                  ))
```

```{r, eval = FALSE}
kras <- condimentsPaper::import_fibrosis()
```

```{r}
data("fibrosis", package = "condimentsPaper")
```

# EDA

Reduced dimension coordinates are obtained from the original publication.

```{r}
df <- bind_cols(reducedDim(sce, "UMAP") %>% as.data.frame(),
                colData(sce)[, 1:14] %>% as.data.frame())
ggplot(df, aes(x = UMAP_1, y = UMAP_2, col = Status)) +
  geom_point(size = .5) +
  scale_color_brewer(palette = "Accent") +
  labs(col = "Disease") +
  theme(legend.position = c(.15, .4),
        legend.background = element_blank()) +
  guides(col = guide_legend(override.aes = list(size = 3)))
ggplot(df, aes(x = UMAP_1, y = UMAP_2, col = celltype)) +
  geom_point(size = .5) +
  scale_color_brewer(palette = "Dark2") +
  labs(col = "Cell Type") +
  theme(legend.position = c(.2, .4),
        legend.background = element_blank()) +
  guides(col = guide_legend(override.aes = list(size = 3)))
```

# Imbalance score

```{r}
scores <- condiments::imbalance_score(
  Object = df %>% select(UMAP_1, UMAP_2) %>% as.matrix(), 
  conditions = df$Status, k = 20, smooth = 30)
df$scores <- scores$scaled_scores
ggplot(df, aes(x = UMAP_1, y = UMAP_2, col = scores)) +
  geom_point(size = .7) +
  scale_color_viridis_c(option = "C", breaks = c(0, 2, 4, 6)) +
  labs(col = "Scores") +
  theme(legend.position = c(.15, .4),
        legend.background = element_blank())
```

# Differential Topology

To estimate the trajectory, we use _slingshot_ [@Street2018a].

## Fit slingshot

```{r}
sce <- slingshot(sce, reducedDim = 'UMAP',
                 clusterLabels = colData(sce)$celltype,
                 start.clus = 'AT1', approx_points = 100)
```

## Topology Test

```{r}
set.seed(821)
topologyTest(SlingshotDataSet(sce), sce$Status)

sce_SCGB3A2 <- slingshot(sce[, sce$celltype != "SCGB3A2+"],
                         reducedDim = 'UMAP',
            clusterLabels = colData(sce[, sce$celltype != "SCGB3A2+"])$celltype,
                 start.clus = 'AT1', approx_points = 100)
set.seed(821)
topologyTest(SlingshotDataSet(sce_SCGB3A2), sce_SCGB3A2$Status)
rm(sce_SCGB3A2)
```

## Separate trajectories 

```{r}
sdss <- slingshot_conditions(SlingshotDataSet(sce), sce$Status)
curves <- bind_rows(lapply(sdss, slingCurves, as.df = TRUE),
                    .id = "Status")

p4a <- ggplot(df, aes(x = UMAP_1, y = UMAP_2, col = Status)) +
  geom_point(size = .7, alpha = .2) +
  scale_color_brewer(palette = "Accent") +
  labs(col = "Disaes") +
  geom_path(data = curves %>% arrange(Status, Lineage, Order),
            aes(group = interaction(Lineage, Status)), size = 1.5) +
  annotate("text", x = -10, y = 6, label = "Lineage1", size = 5) +
  annotate("text", x = -7, y = -2.7, label = "Lineage2", size = 5) +
  theme(legend.position = c(.15, .35),
        legend.background = element_blank()) +
  NULL
p4a
```

## Common trajectory

```{r}
df <- bind_cols(
  as.data.frame(reducedDim(sce, "UMAP")),
  slingPseudotime(sce) %>% as.data.frame() %>%
    dplyr::rename_with(paste0, "_pst", .cols = everything()),
  slingCurveWeights(sce) %>% as.data.frame(),
  ) %>%
  mutate(Lineage1_pst = if_else(is.na(Lineage1_pst), 0, Lineage1_pst),
         Lineage2_pst = if_else(is.na(Lineage2_pst), 0, Lineage2_pst),
         pst = if_else(Lineage1 > Lineage2, Lineage1_pst, Lineage2_pst),
         pst = max(pst) - pst)
curves <- slingCurves(sce, as.df = TRUE)
ggplot(df, aes(x = UMAP_1, y = UMAP_2)) +
  geom_point(size = .7, aes(col = pst)) +
  scale_color_viridis_c() +
  labs(col = "Pseudotime") +
  geom_path(data = curves %>% arrange(Order),
            aes(group = Lineage), col = "black", size = 1.5) +
  annotate("text", x = -10, y = 6, label = "Lineage1", size = 5) +
  annotate("text", x = -7, y = -2.7, label = "Lineage2", size = 5) +
  theme(legend.position = c(.15, .35),
        legend.background = element_blank()) 
```

# Differential Progression
## Test

```{r}
progressionTest(sce, conditions = sce$Status, lineages = TRUE)
```

## Plot

```{r}
df <-  slingPseudotime(sce) %>% as.data.frame() 
df$Status <- sce$Status
df <- df %>% 
  pivot_longer(-Status, names_to = "Lineage",
               values_to = "pst") %>%
  filter(!is.na(pst))
ggplot(df, aes(x = pst)) +
  geom_density(alpha = .8, aes(fill = Status), col = "transparent") +
  geom_density(aes(col = Status), fill = "transparent", size = 1.5) +
  labs(x = "Pseudotime", fill = "Status") +
  facet_wrap(~Lineage, scales = "free_x") +
  guides(col = "none", fill = guide_legend(
    override.aes = list(size = 1.5, col = c("#7FC97F", "#BEAED4"))
  )) +
  scale_fill_brewer(palette = "Accent") +
  scale_color_brewer(palette = "Accent") +
  theme(legend.position = c(.8, .8), legend.direction = "horizontal")
```

# Differential fate selection
## Test

```{r}
fateSelectionTest(sce, conditions = sce$Status)
```

## Plot

```{r}
df <- slingCurveWeights(SlingshotDataSet(sce), as.probs = TRUE) %>% as.data.frame()
df$Status <- sce$Status
df <- df %>%
  pivot_longer(-Status, names_to = "Lineage", values_to = "ws")
ggplot(df, aes(x = ws)) +
  geom_density(alpha = .8, aes(fill = Status), col = "transparent") +
  geom_density(aes(col = Status), fill = "transparent", size = 1.5) +
  labs(x = "Weights", fill = "Status") +
  facet_wrap(~Lineage, scales = "free_x") +
  guides(fill = "none", col = "none") +
  scale_fill_brewer(palette = "Accent") +
  scale_color_brewer(palette = "Accent")
```

# Differential expression

We use _tradeSeq_ [@VandenBerge2020].

## Select number of knots

```{r evaluateK, eval = FALSE}
library(tradeSeq)
set.seed(3)
BPPARAM <- BiocParallel::bpparam()
BPPARAM$workers <- 3
icMat <- evaluateK(counts = sce, sds = SlingshotDataSet(sce), 
                   conditions = factor(sce$Status),
                   nGenes = 300, parallel = TRUE, BPPARAM = BPPARAM, k = 3:7)
```

### Fit GAM

Next, we fit the NB-GAMs using 6 knots, based on the pseudotime and cell-level weights estimated by Slingshot. We use the `conditions` argument to fit separate smoothers for each condition.

```{r, eval = FALSE}
library(tradeSeq)
set.seed(3)
sce <- fitGAM(counts = sce, conditions = factor(sce$Status), nknots = 6)
```

## Differential expression between conditions

```{r, eval=TRUE}
condRes <- conditionTest(sce, l2fc = log2(2), lineages = TRUE)
condRes$padj <- p.adjust(condRes$pvalue_lineage1, "fdr")
mean(condRes$padj <= 0.05, na.rm = TRUE)
sum(condRes$padj <= 0.05, na.rm = TRUE)
conditionGenes <- rownames(condRes)[condRes$padj <= 0.05]
conditionGenes <- conditionGenes[!is.na(conditionGenes)]
```

## Differential expression between lineages

```{r}
patRes <- patternTest(sce, l2fc = log2(2))
patRes$padj <- p.adjust(patRes$pvalue, "fdr")
mean(patRes$padj <= 0.05, na.rm = TRUE)
sum(patRes$padj <= 0.05, na.rm = TRUE)
patternGenes <- rownames(patRes)[patRes$padj <= 0.05]
patternGenes <- patternGenes[!is.na(patternGenes)]
```

### Visualize select genes from the og paper

```{r, eval=TRUE}
library(RColorBrewer)
scales <- c("#FCAE91", "#DE2D26", "#BDD7E7", "#3182BD")

# plot genes
oo <- order(patRes$waldStat, decreasing = TRUE)
genes <- c("SCGB3A2", "SFTPC", "ABCA3", "AGER")

# most significant gene
ps <- lapply(genes, function(gene){
  plotSmoothers(sce, assays(sce)$counts, gene = gene, curvesCol = scales,
                border = FALSE, sample = .2) +
    scale_x_reverse(breaks = 0:4/.4, labels = 4:0/.4) +
    labs(title = gene, col = "Lineage and Condition") +
    scale_color_manual(values = scales) +
    theme(plot.title = element_text(hjust = .5),
          rect = element_blank())
  }
)
legend_gene <- get_legend(
  ps[[1]] + theme(legend.position = "bottom") +
    guides(col = guide_legend(title.position = "top"))
)
ps <- lapply(ps, function(p) {p + guides(col = "none")})
p6 <- plot_grid(
  plot_grid(plotlist = ps, ncol = 2, scale = .95,
            labels = c('a)', 'b)', 'c)', 'd)')),
  plot_grid(NULL, legend_gene, NULL, rel_widths = c(1, 1, 1), ncol = 3), 
  nrow = 2, rel_heights = c(4, .5)
)
```

### Heatmaps of genes DE between lineages

Below we show heatmaps of the genes DE between lineages The DE genes in the heatmaps are ordered according to a hierarchical clustering on the first lineage.

```{r, eval=TRUE}
### based on mean smoother
yhatSmooth <- predictSmooth(sce, gene = patternGenes, nPoints = 50, tidy = TRUE) %>%
  mutate(yhat = log1p(yhat)) %>%
  group_by(gene) %>%
  mutate(yhat = scales::rescale(yhat)) %>%
  filter(condition == "ILD") %>%
  select(-condition) %>%
  ungroup()
heatSmooth_Lineage1 <- pheatmap(
  yhatSmooth %>%
    filter(lineage == 1) %>%
    select(-lineage) %>%
    arrange(-time) %>%
    pivot_wider(names_from = time, values_from = yhat) %>%
    select(-gene),
  cluster_cols = FALSE, show_rownames = FALSE, show_colnames = FALSE,
  main = "Lineage 1", legend = FALSE, silent = TRUE
)

heatSmooth_Lineage2 <- pheatmap(
  (yhatSmooth %>%
    filter(lineage == 2) %>%
    select(-lineage) %>%
    arrange(-time) %>%
    pivot_wider(names_from = time, values_from = yhat) %>%
    select(-gene))[heatSmooth_Lineage1$tree_row$order, ],
  cluster_cols = FALSE, cluster_rows = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "Lineage 2",
  legend = FALSE, silent = TRUE 
)

p7 <- plot_grid(heatSmooth_Lineage1[[4]], heatSmooth_Lineage2[[4]], ncol = 2)
p7
```

# Session info 

```{r}
sessionInfo()
```

# References
