---
title: 'Analysis of the KRAS datataset'
author: "Hector Roux de Bézieux"
bibliography: ref.bib
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Analysis of the KRAS dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r packages, include=F}
library(knitr)
opts_chunk$set(
  fig.pos = "!h", out.extra = "", warning = F, 
  fig.align = "center"
)
```

# Load data

The dataset we will be working with concerns a single-cell RNA-sequencing dataset consisting of three cancer models under a  KRAS(G12C) inhibition [@Xue2020].

```{r}
libs <- c("here", "dplyr", "tradeSeq", "SingleCellExperiment", "slingshot",
           "condiments", "scater", "RColorBrewer", "pheatmap", "cowplot",
          "tidyr")
suppressMessages(
  suppressWarnings(sapply(libs, require, character.only = TRUE))
)
rm(libs)
theme_set(theme_classic())
```

```{r, eval = FALSE}
kras <- condimentsPaper::import_KRAS()
```

```{r}
data("kras", package = "condimentsPaper")
```

# EDA

Reduced dimension coordinates are obtained from the original publication.

```{r}
cols <- c(brewer.pal(5, "Blues")[2:5], brewer.pal(3, "Greens")[2:3], brewer.pal(3, "Reds")[2:3],
          brewer.pal(3, "Oranges")[2], "Grey")
names(cols) <- c(3, 5, 4, 1, 8, 2, 9, 10, 6, 7)
kras$Cluster <- as.character(kras$Cluster)
reducedDim(kras, "TSNE") <- colData(kras)[, c("tSNE1", "tSNE2")] %>% as.matrix()
df <- colData(kras) %>% as.data.frame() %>%
  sample_frac(1)
p1 <- ggplot(df, aes(x = tSNE1, y = tSNE2, col = Batch)) +
  geom_point(size = .7) +
  scale_color_brewer(palette = "Accent") +
  labs(col = "Type")
p1
p2 <- ggplot(df, aes(x = tSNE1, y = tSNE2, fill = Cluster)) +
  geom_point(size = 1, alpha = .65, col = "grey70", shape = 21) +
  scale_fill_manual(values = cols) +
  labs(fill = "Cell Clusters")
p2
```

# Imbalance score

```{r}
kras <- imbalance_score(Object = kras, conditions = "Batch", dimred = "TSNE")
df$scores <- kras[, df$X1]$scores$scaled_scores
p3 <- ggplot(df, aes(x = tSNE1, y = tSNE2, col = scores)) +
  geom_point(size = .7) +
  scale_color_viridis_c(option = "C") +
  labs(col = "Score")
p3
```

# Differential Topology

To estimate the trajectory, we use _slingshot_ [@Street2018a].

## Fit slingshot

```{r}
kras <- slingshot(kras, reducedDim = "TSNE", clusterLabels = kras$Cluster,
                 start.clus = 7, extend = "n", reweight = FALSE, reassign = FALSE)
```

## Topology Test

```{r}
topologyTest(kras, conditions = "Batch", rep = 50)
```

## Individual curves

```{r}
sdss <- slingshot_conditions(kras, kras$Batch, approx_points = FALSE,
                             extend = "n", reweight = FALSE, reassign = FALSE)
sdss$condition_id <- names(sdss)
sdss$mapping <- matrix(rep(1:3, each = 3), nrow = 3, ncol = 3, byrow = TRUE)
sds <- do.call(merge_sds, sdss)
```

```{r}
df <- df %>% select(X1, tSNE1, tSNE2, Cluster, Batch) %>%
  full_join(
    slingPseudotime(sds) %>% 
      as.data.frame() %>%
      mutate(X1 = rownames(.)) %>%
      dplyr::rename("Lineage1" = V1, "Lineage2" = V2, "Lineage3" = V3))
df <- df %>%
  pivot_longer(starts_with("Lineage"), names_to = "Curve", values_to = "pst")
p4 <- ggplot(df, aes(x = tSNE1, y = tSNE2, col = Batch)) +
  geom_point(size = .7, alpha = .1) +
  scale_color_brewer(palette = "Accent")
for (batch in unique(kras$Batch)) {
  sds_cond <- sdss[[batch]]
  for (i in 1:3) {
    p4 <- p4 +  
      geom_path(data = slingCurves(sds_cond)[[i]]$s[slingCurves(sds_cond)[[i]]$ord, ] %>%
                  as.data.frame() %>%
                  mutate(Batch = batch), 
                size = 1.5)   
  }
}
position <- data.frame(
  "tSNE1" = c(40, -30, 45),
  "tSNE2" = c(50, -50, -50),
  "Batch" = "H2122A",
  "text" = paste0("Lineage ", 1:3)
)
p4 <- p4 + 
  geom_text(data = position, col = "black", aes(label = text), size = 5)
p4
```

# Differential Progression
## Test

```{r}
progressionTest(sds, conditions = kras$Batch, lineages = TRUE)
```

## Plot

```{r}
p5 <- ggplot(df, aes(x = pst)) +
  geom_density(alpha = .4, aes(fill = Batch), col = "transparent") +
  geom_density(aes(col = Batch), fill = "transparent", size = 1.5) +
  guides(col = FALSE) +
  scale_fill_brewer(palette = "Accent") +
  scale_color_brewer(palette = "Accent") +
  labs(x = "Pseudotime", fill = "Type") +
  facet_wrap(~ Curve, scales = "free_x")
p5
```

# Differential Differentiation
## Test

```{r}
differentiationTest(sds, conditions = kras$Batch, pairwise = TRUE)
```

## Plot

```{r}
weights <- condiments:::.sling_reassign(sds)
df <- df %>%
  full_join(weights %>% 
              as.data.frame() %>%
              mutate(X1 = rownames(.)) %>%
              dplyr::rename("Lineage1" = V1, "Lineage2" = V2, "Lineage3" = V3) %>%
              pivot_longer(starts_with("Lineage"), names_to = "Curve", values_to = "weights")
      )
df_w <- df %>%
  select(-pst) %>%
  group_by(X1) %>%
  mutate(weights = weights / sum(weights)) %>%
  pivot_wider(names_from = "Curve", values_from = "weights")
p <- ggplot(df_w, aes(x = Lineage1, y = Lineage3)) +
  # geom_hex(aes(fill=log(..count..))) +
  geom_hex() +
  scale_fill_viridis_c(direction = -1) +
  facet_wrap(~Batch, scales = "free") +
  geom_abline(slope = -1, intercept = 1, linetype = "dotted") +
  geom_abline(slope = -1, intercept = 2/3, linetype = "dotted") +
  geom_abline(slope = -1, intercept = 1/3, linetype = "dotted") +
  annotate("text", x = .53, y = .53, label = "w3 = 0", angle = -52) +
  annotate("text", x = .62, y = .1, label = "w3 = 1/3", angle = -52) +
  annotate("text", x = .14, y = .14, label = "w3 = 2/3", angle = -52) +
  theme(legend.position = "bottom") +
  labs(x = "Weights for Lineage 1 (w1)", y = "Weights for Lineage 2 (w2)",
       fill = "counts per hexagon")
p
```

```{r}
df_w <- df %>%
  select(-pst) %>%
  group_by(X1) %>%
  mutate(weights = weights / sum(weights)) %>%
  ungroup() %>%
  group_by(Batch, Curve) %>%
  summarise(weights = mean(weights), .groups = NULL)
p2 <- ggplot(df_w, aes(x = Curve, fill = Batch, y = weights)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Accent") +
  theme(legend.position = c(.7, .7)) +
  labs(x = "", y = "Mean weight")
p2
```

# Differential expression

We use _tradeSeq_ [@VandenBerge2020].

```{r}
set.seed(3)
filter <- apply(counts(kras), 1, function(g) {
    sum(g >= 5) >= 10
})
kras <- kras[filter, ]
```

### Select number of knots

```{r evaluateK, eval = FALSE}
set.seed(3)
library(BiocParallel)
BPPARAM <- BiocParallel::bpparam()
BPPARAM$workers <- 4
icMat <- evaluateK(counts = as.matrix(assays(kras)$counts),
                   pseudotime = slingPseudotime(sds, na = FALSE),
                   cellWeights = weights,
                   conditions = factor(colData(kras)$Batch),
                   nGenes = 300,
                   k = 3:7,
                   parallel = TRUE,
                   BPPARAM = BPPARAM)
```

### Fit GAM

Next, we fit the NB-GAMs using 5 knots, based on the pseudotime and cell-level weights estimated by Slingshot. We use the `conditions` argument to fit separate smoothers for each condition.

```{r, eval = FALSE}
set.seed(3)
kras@int_metadata$slingshot <- sds
kras <- fitGAM(counts = kras,
               conditions = factor(colData(kras)$Batch),
               parallel = TRUE,
               genes = 1:10,
               BPPARAM = BPPARAM,
               nknots = 6)
```

## Differential expression between conditions

```{r, eval=TRUE}
condRes <- conditionTest(kras, l2fc = log2(2), lineages = TRUE)
condRes$padj <- p.adjust(condRes$pvalue, "fdr")
mean(condRes$padj <= 0.05, na.rm = TRUE)
sum(condRes$padj <= 0.05, na.rm = TRUE)
conditionGenes <- rownames(condRes)[condRes$padj <= 0.05]
conditionGenes <- conditionGenes[!is.na(conditionGenes)]
```

## Visualize most and least significant gene

```{r, eval=TRUE}
# plot genes
scales <- c("#7FC97F", "#57A059", "#2E7935",
            "#BEAED4", "#9687AC", "#706285",
            "#FDC086", "#CC945C", "#9D6A34")
scales <- scales[c(1, 4, 7, 2, 5, 8, 3, 6, 9)]
oo <- order(condRes$waldStat, decreasing = TRUE)

# most significant gene
p6 <- plotSmoothers(kras, assays(kras)$counts,
              gene = rownames(assays(kras)$counts)[oo[1]],
              alpha = 1, curvesCols = scales, sample = .3) +
  scale_color_manual(values = scales) +
  ggtitle(rownames(assays(kras)$counts)[oo[1]])

# Second most significant gene
p7 <- plotSmoothers(kras, assays(kras)$counts,
              gene = rownames(assays(kras)$counts)[oo[2]],
              alpha = 1, curvesCols = scales, sample = .3) +
  scale_color_manual(values = scales) +
  ggtitle(rownames(assays(kras)$counts)[oo[2]])

# least significant gene
p8 <- plotSmoothers(kras, assays(kras)$counts,
              gene = rownames(assays(kras)$counts)[oo[nrow(kras)]],
              alpha = 1, curvesCols = scales, sample = .3) +
  scale_color_manual(values = scales) +
  ggtitle(rownames(assays(kras)$counts)[oo[nrow(kras)]])
p6
p7
p8
```

## Heatmaps of genes DE between conditions for lineage 1

Below we show heatmaps of the genes DE between conditions. The DE genes in the heatmaps are ordered according to a hierarchical clustering on the TGF-Beta condition.

```{r, eval=TRUE}
### based on mean smoother
condRes$padj_lineage1 <- p.adjust(condRes$pvalue_lineage1, "fdr")
conditionGenes_lineage1 <- rownames(condRes)[condRes$padj_lineage1 <= 0.05]
conditionGenes_lineage1 <- conditionGenes_lineage1[!is.na(conditionGenes_lineage1)]

yhatSmooth <- predictSmooth(kras, gene = conditionGenes_lineage1, nPoints = 50, tidy = FALSE) %>%
  log1p()
yhatSmoothScaled <- t(apply(yhatSmooth[, c(1:50, 151:200, 301:350)], 1, scales::rescale))
heatSmooth_H2122A <- pheatmap(yhatSmoothScaled[, 1:50],
  cluster_cols = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "H2122A", legend = FALSE,
  silent = TRUE
)

matchingHeatmap_H358A <- pheatmap(yhatSmoothScaled[heatSmooth_H2122A$tree_row$order, 51:100],
  cluster_cols = FALSE, cluster_rows = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "H358A",
  legend = FALSE, silent = TRUE
)

matchingHeatmap_SW1573A <- pheatmap(yhatSmoothScaled[heatSmooth_H2122A$tree_row$order, 101:150],
  cluster_cols = FALSE, cluster_rows = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "SW1573A",
  legend = FALSE, silent = TRUE
)

p9 <- plot_grid(heatSmooth_H2122A[[4]], matchingHeatmap_H358A[[4]], matchingHeatmap_SW1573A[[4]],
                NULL, NULL, NULL, ncol = 3, rel_widths = c(1.4, 1, 1), rel_heights = c(10, 1)) +
  draw_text("Lineage 1", x = .5, y = .05)
p9
```

## Heatmaps of genes DE between conditions for lineage 2

Below we show heatmaps of the genes DE between conditions. The DE genes in the heatmaps are ordered according to a hierarchical clustering on the TGF-Beta condition.

```{r, eval=TRUE}
condRes$padj_lineage2 <- p.adjust(condRes$pvalue_lineage2, "fdr")
conditionGenes_lineage2 <- rownames(condRes)[condRes$padj_lineage2 <= 0.05]
conditionGenes_lineage2 <- conditionGenes_lineage1[!is.na(conditionGenes_lineage2)]

yhatSmooth <- predictSmooth(kras, gene = conditionGenes_lineage2, nPoints = 50, tidy = FALSE) %>%
  log1p()
yhatSmoothScaled <- t(apply(yhatSmooth[, c(51:100, 201:250, 351:400)],1, scales::rescale))

heatSmooth_H2122A <- pheatmap(yhatSmoothScaled[, 1:50],
  cluster_cols = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "H2122A", legend = FALSE,
  silent = TRUE
)

matchingHeatmap_H358A <- pheatmap(yhatSmoothScaled[heatSmooth_H2122A$tree_row$order, 51:100],
  cluster_cols = FALSE, cluster_rows = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "H358A",
  legend = FALSE, silent = TRUE
)

matchingHeatmap_SW1573A <- pheatmap(yhatSmoothScaled[heatSmooth_H2122A$tree_row$order, 101:150],
  cluster_cols = FALSE, cluster_rows = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "SW1573A",
  legend = FALSE, silent = TRUE
)

p10 <- plot_grid(heatSmooth_H2122A[[4]], matchingHeatmap_H358A[[4]], matchingHeatmap_SW1573A[[4]],
                NULL, NULL, NULL, ncol = 3, rel_widths = c(1.4, 1, 1), rel_heights = c(10, 1)) +
  draw_text("Lineage 2", x = .5, y = .05)
p10
```

## Heatmaps of genes DE between conditions for lineage 3

```{r, eval=TRUE}
condRes$padj_lineage3 <- p.adjust(condRes$pvalue_lineage3, "fdr")
conditionGenes_lineage3 <- rownames(condRes)[condRes$padj_lineage3 <= 0.05]
conditionGenes_lineage3 <- conditionGenes_lineage3[!is.na(conditionGenes_lineage3)]

yhatSmooth <- predictSmooth(kras, gene = conditionGenes_lineage3, nPoints = 50, tidy = FALSE) %>%
  log1p()
yhatSmoothScaled <- t(apply(yhatSmooth[, c(101:150, 251:300, 401:450)],1, scales::rescale))
heatSmooth_H2122A <- pheatmap(yhatSmoothScaled[, 1:50],
  cluster_cols = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "H2122A", legend = FALSE,
  silent = TRUE
)

matchingHeatmap_H358A <- pheatmap(yhatSmoothScaled[heatSmooth_H2122A$tree_row$order, 51:100],
  cluster_cols = FALSE, cluster_rows = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "H358A",
  legend = FALSE, silent = TRUE
)

matchingHeatmap_SW1573A <- pheatmap(yhatSmoothScaled[heatSmooth_H2122A$tree_row$order, 101:150],
  cluster_cols = FALSE, cluster_rows = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "SW1573A",
  legend = FALSE, silent = TRUE
)

p11 <- plot_grid(heatSmooth_H2122A[[4]], matchingHeatmap_H358A[[4]], matchingHeatmap_SW1573A[[4]],
                 NULL, NULL, NULL, ncol = 3, rel_widths = c(1.4, 1, 1), rel_heights = c(10, 1)) +
  draw_text("Lineage 3", x = .5, y = .05)
p11
```

# Session info 

```{r}
sessionInfo()
```

# References
