---
title: 'Running and analyzing the simulations'
author: "Hector Roux de Bézieux"
bibliography: ref.bib
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Running and analyzing the simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r packages, include=F}
library(knitr)
opts_chunk$set(
  fig.pos = "!h", out.extra = "", warning = F, 
  fig.align = "center"
)
```


```{r}
libs <- c("here", "dplyr", "stringr", "tidyr", "SingleCellExperiment", "slingshot",
          "condiments", "cowplot", "dyngen", "ggplot2")
suppressMessages(
  suppressWarnings(sapply(libs, require, character.only = TRUE))
)
rm(libs)
theme_set(theme_classic())
```

The simulations are based on the _dyngen_ package[@Cannoodt2020].

# Running the simulations

This can be very long, we recommand running this in a separate script. None of the code in this section is run while compiling the vignettes.

## Setting the parameters

```{r, eval = FALSE}
library(BiocParallel)
set.seed(97174)
options(Ncpus = 16L)
BPPARAM <- bpparam()
BPPARAM$progressbar <- TRUE
BPPARAM$workers <- 30
ratio <- c(.5, .8, .9, .95)
n_boot_real <- 50
frac_real <- .1
n_boot_null <- 100
frac_null <- .05
```

```{r}
sub_sample <- function(sce, frac) {
  df <- colData(sce) %>% as.data.frame() %>%
    mutate(id = as.numeric(rownames(.))) %>%
    group_by(condition) %>%
    sample_frac(frac)
  return(sce[, df$id])
}
```

## Bifurcating trajectory with two conditions

```{r, eval = FALSE}
res <- list()
for (multiplier in c(ratio, round(1 / ratio, 2)[4:1])) {
  sce <- create_bifurcating_simu(multiplier = multiplier)
  if (multiplier == .5) fork_sce <- sce
  set.seed(2019)
  res_loc <- bplapply(seq_len(n_boot_real), function(b, frac) {
    return(anayze_all(.sub_sample(sce, frac = frac)))
  }, frac = frac_real, BPPARAM = BPPARAM)
  res[[as.character(multiplier)]] <- bind_rows(res_loc)
}
real <- bind_rows(res, .id = "multiplier") %>%
  clean_results()
```

```{r, eval = FALSE}
sce <- create_bifurcating_simu(multiplier = 1, nSim = 400)
set.seed(2019)
res_loc <- bplapply(seq_len(n_boot_null), function(b, frac) {
  return(anayze_all(sub_sample(sce, frac = frac)))
}, frac = frac_null, BPPARAM = BPPARAM)
null <- bind_rows(res_loc) %>%
  mutate(multiplier = 1) %>%
  clean_results()
```

```{r, eval = FALSE}
fork <- bind_rows("Real" = real, "Null" = null, .id = "Type") %>%
  arrange(multiplier)
```

## Two bifurcations and two conditions

```{r, eval = FALSE}
res <- list()
for (multiplier in c(ratio, round(1 / ratio, 2)[4:1])) {
 sce <- create_consecutive_bifurcating_simu(multiplier = multiplier)
 if (multiplier == .5) tree_sce <- sce
  set.seed(2019)
  res_loc <- bplapply(seq_len(n_boot_real), function(b, frac) {
    return(anayze_all(.sub_sample(sce, frac = frac)))
  }, frac = frac_real, BPPARAM = BPPARAM)
  res[[as.character(multiplier)]] <- bind_rows(res_loc)
}
real <- bind_rows(res, .id = "multiplier") %>%
  clean_results()
```

```{r, eval = FALSE}
sce <- create_consecutive_bifurcating_simu(multiplier = 1, nSim = 400)
set.seed(2019)
res_loc <- bplapply(seq_len(n_boot_null), function(b, frac) {
  return(anayze_all(sub_sample(sce, frac = frac)))
}, frac = frac_null, BPPARAM = BPPARAM)
null <- bind_rows(res_loc) %>%
  mutate(multiplier = 1) %>%
  clean_results()
```

```{r, eval = FALSE}
tree <- bind_rows("Real" = real, "Null" = null, .id = "Type") %>%
  arrange(multiplier)
```

## Bifurcating trajectory with three conditions

```{r, eval = FALSE}
res <- list()
for (multiplier in ratio) {
 sce <- create_bifurcating_three_conditions_simu(multiplier = multiplier)
 if (multiplier == .5) complex_sce <- sce
  set.seed(2019)
  res_loc <- bplapply(seq_len(n_boot_real), function(b, frac) {
    return(anayze_all(.sub_sample(sce, frac = frac)))
  }, frac = frac_real, BPPARAM = BPPARAM)
  res[[as.character(multiplier)]] <- bind_rows(res_loc)
}
real <- bind_rows(res, .id = "multiplier") %>%
  clean_results()
```

```{r, eval = FALSE}
sce <- create_bifurcating_three_conditions_simu(multiplier = 1, nSim = 400)
set.seed(2019)
res_loc <- bplapply(seq_len(n_boot_null), function(b, frac) {
  return(anayze_all(sub_sample(sce, frac = frac)))
}, frac = frac_null, BPPARAM = BPPARAM)
null <- bind_rows(res_loc) %>%
  mutate(multiplier = 1) %>%
  clean_results()
```

```{r, eval = FALSE}
complex <- bind_rows("Real" = real, "Null" = null, .id = "Type") %>%
  arrange(multiplier)
```

# Exploring the simulations
## Low dimension representation
### Bifurcating trajectory with two conditions

```{r, fig.height=5, fig.width=6}
data("fork_sce", package = "condimentsPaper")
plot_reduced_dim_together(fork_sce)
```

```{r, fig.height=5, fig.width=6}
data("tree_sce", package = "condimentsPaper")
plot_reduced_dim_together(tree_sce)
```

```{r, fig.height=5, fig.width=6}
data("complex_sce", package = "condimentsPaper")
plot_reduced_dim_together(complex_sce)
```

## Network branching

```{r}
set.seed(0989021)
backbone <- backbone_bifurcating()
model_common <-
  initialise_model(
    backbone = backbone,
    num_cells = 100,
    num_tfs = nrow(backbone$module_info),
    num_targets = 250,
    num_hks = 250,
    simulation_params = simulation_default(
      census_interval = 10, 
      ssa_algorithm = ssa_etl(tau = 300 / 3600),
      experiment_params = simulation_type_wild_type(num_simulations = 100)
    )
  ) %>%
  generate_tf_network()
set.seed(298)
net <- plot_backbone_modulenet_simplify(model_common) +
  guides(col = FALSE, edge_width = FALSE) +
  ggraph::scale_edge_width_continuous(range = c(.5, .5)) + 
  theme(plot.background = element_blank()) 
net <- ggdraw() +
  draw_plot(net, 0, 0, scale = 1.2)
net
```

```{r}
set.seed(1082)
backbone <- backbone_consecutive_bifurcating()
model_common <-
  initialise_model(
    backbone = backbone,
    num_cells = 100,
    num_tfs = nrow(backbone$module_info),
    num_targets = 250,
    num_hks = 250,
    simulation_params = simulation_default(
      census_interval = 10, 
      ssa_algorithm = ssa_etl(tau = 300 / 3600),
      experiment_params = simulation_type_wild_type(num_simulations = 100)
    )
  ) %>%
  generate_tf_network()
set.seed(2)
net2 <- plot_backbone_modulenet_simplify(model_common) +
  guides(col = FALSE, edge_width = FALSE) +
  ggraph::scale_edge_width_continuous(range = c(.5, .5)) + 
  theme(plot.background = element_blank()) 
net2 <- ggdraw() +
  draw_plot(net2, 0, 0, scale = 1.2)
net2
```

# Analyzing the simulations

```{r}
data("fork", package = "condimentsPaper")
data("tree", package = "condimentsPaper")
data("complex", package = "condimentsPaper")
```

## All together

```{r}
df <- lapply(list("fork" = fork, "tree" = tree, "complex" = complex), 
             function(dataset) {
   bind_rows(".05" = all_metrics(dataset, .05),
             ".01" = all_metrics(dataset, .01), 
             .id = "cutoff") %>%
    return()
             }
) %>% bind_rows(.id = "dataset") %>%
  mutate(value = round(value, 2))
```

```{r, fig.height=6, fig.width=6}
labs_pct <- function(Panel) {
  val <- word(Panel) %>% as.numeric()
  return(paste0("FDR: ", round(100 * val), "%"))
}
df <- df %>%
  mutate(metric = factor(metric, levels = c("TNR", "PPV", "TPR", "NPV", "F1")),
         dataset = factor(dataset, levels = c("fork", "tree", "complex"))) %>%
  arrange(dataset, metric) %>%
  mutate(test_type = factor(test_type, 
                            levels = c("condiments_prog", "condiments_diff",
                                       "DAseq", "milo")),
         Panel = paste0(cutoff, " ", dataset),
         Panel = factor(Panel, levels = unique(Panel)),
         label = if_else(is.na(value), "NA", as.character(value)))
p <- ggplot(df, aes(x = test_type, y = metric, label = label,
               fill = scaled)) +
  geom_tile() +
  geom_text(na.rm = FALSE) +
  scale_fill_viridis_c() +
  scale_x_discrete(labels = c("progressionTest", "differentationTest", "DAseq",
                              "milo")) +
  labs(x = "", y = "", fill = "Difference with\noptimal value") +
  theme(legend.position = "bottom") +
  facet_grid(Panel ~ ., switch = "y", labeller = labeller(Panel = labs_pct)) +
  theme(axis.text.x = element_text(size = 14, angle = -20, vjust = .2))
p
```

## Special look at three condition situation

```{r}
ggplot(complex, aes(x = as.numeric(factor(multiplier, levels = unique(multiplier))),
                    y = adjusted_p_value)) +
  geom_boxplot(aes(group = interaction(test_type, multiplier),
                   fill = test_type)) +
  geom_hline(yintercept = .05) +
  scale_x_continuous(labels = unique(complex$multiplier),
                     breaks = 1:5) +
  scale_color_brewer(palette = "Pastel2") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(x = "Effect Size: Multiplier on the master regulator",
       y = "adjusted p-value", col = "Test", fill = "Test") +
  theme(legend.position = "bottom")
```
