---
title: 'Running and analyzing the simulations'
author: "Hector Roux de Bézieux"
bibliography: ref.bib
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Running and analyzing the simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r packages, include=F}
library(knitr)
opts_chunk$set(
  fig.pos = "!h", out.extra = "", warning = F, 
  fig.align = "center"
)
```


```{r}
libs <- c("here", "dplyr", "stringr", "tidyr", "SingleCellExperiment", "slingshot",
          "condiments", "cowplot", "dyngen", "ggplot2", "condimentsPaper")
suppressMessages(
  suppressWarnings(sapply(libs, require, character.only = TRUE))
)
rm(libs)
theme_set(theme_classic())
```

The simulations are based on the _dyngen_ package[@Cannoodt2020].

# Running the simulations

This can be very long, we recommand running this in a separate script. None of the code in this section is run while compiling the vignettes.

## Setting the parameters

```{r, eval = FALSE}
library(BiocParallel)
set.seed(97174)
options(Ncpus = 16L)
nCores <- 16
ratio <- c(.5, .8, .9, .95)
n_boot_real <- 50
frac_real <- .1
n_boot_null <- 100
frac_null <- .05
```

```{r}
sub_sample <- function(sce, frac) {
  df <- colData(sce) %>% as.data.frame() %>%
    mutate(id = as.numeric(rownames(.))) %>%
    group_by(condition) %>%
    sample_frac(frac)
  return(sce[, df$id])
}
```

## Bifurcating trajectory with two conditions

```{r, eval = FALSE}
res <- list()
for (multiplier in c(ratio, round(1 / ratio, 2)[4:1])) {
  sce <- create_bifurcating_simu(multiplier = multiplier)
  if (multiplier == .5) fork_sce <- sce
  set.seed(2019)
  res_loc <- mclapply(seq_len(n_boot_real), function(b, frac) {
      return(anayze_all(sub_sample(sce, frac = frac)))
    }, frac = frac_real, mc.cores = nCores)
    worked <- lapply(res_loc, function(res){
      return(!"SingleCellExperiment" %in% class(res))
    }) %>% unlist()
    res_loc <- res_loc[worked]
    res[[as.character(multiplier)]] <- bind_rows(res_loc)
}
real <- bind_rows(res, .id = "multiplier") %>%
  clean_results()
```

```{r, eval = FALSE}
sce <- create_bifurcating_simu(multiplier = 1, nSim = 400)
set.seed(2019)
res_loc <- mclapply(seq_len(n_boot_null), function(b, frac) {
  return(anayze_all(sub_sample(sce, frac = frac)))
  }, frac = frac_null, mc.cores = nCores)
worked <- lapply(res_loc, function(res){
  return(!"SingleCellExperiment" %in% class(res))
}) %>% unlist()
res_loc <- res_loc[worked]
null <- bind_rows(res_loc) %>%
  mutate(multiplier = 1) %>%
  clean_results()
```

```{r, eval = FALSE}
fork <- bind_rows("Real" = real, "Null" = null, .id = "Type") %>%
  arrange(multiplier)
```

## Two bifurcations and two conditions

```{r, eval = FALSE}
res <- list()
for (multiplier in c(ratio, round(1 / ratio, 2)[4:1])) {
 sce <- create_consecutive_bifurcating_simu(multiplier = multiplier)
 if (multiplier == .5) tree_sce <- sce
  set.seed(2019)
  res_loc <- mclapply(seq_len(n_boot_real), function(b, frac) {
    return(anayze_all(sub_sample(sce, frac = frac), shape = 3))
  }, frac = frac_real, mc.cores = nCores)
  worked <- lapply(res_loc, class) %>% unlist()
  worked <- worked == "data.frame"
  res_loc <- res_loc[worked]
  res[[as.character(multiplier)]] <- bind_rows(res_loc)
}
real <- bind_rows(res, .id = "multiplier") %>%
  clean_results()
```

```{r, eval = FALSE}
sce <- create_consecutive_bifurcating_simu(multiplier = 1, nSim = 400)
set.seed(2019)
res_loc <- mclapply(seq_len(n_boot_null), function(b, frac) {
  return(anayze_all(sub_sample(sce, frac = frac), shape = 3))
}, frac = frac_null, mc.cores = nCores)
worked <- lapply(res_loc, class) %>% unlist()
worked <- worked == "data.frame"
res_loc <- res_loc[worked]
null <- bind_rows(res_loc) %>%
  mutate(multiplier = 1) %>%
  clean_results()
```

```{r, eval = FALSE}
tree <- bind_rows("Real" = real, "Null" = null, .id = "Type") %>%
  arrange(multiplier)
```

## Bifurcating trajectory with three conditions

```{r, eval = FALSE}
res <- list()
for (multiplier in ratio) {
 sce <- create_bifurcating_three_conditions_simu(multiplier = multiplier)
 if (multiplier == .5) complex_sce <- sce
  set.seed(2019)
  res_loc <- mclapply(seq_len(n_boot_real), function(b, frac) {
    return(anayze_multiple_conditions(sub_sample(sce, frac = frac)))
  }, frac = frac_real, mc.cores = nCores)
  worked <- lapply(res_loc, function(res){
    return(!"SingleCellExperiment" %in% class(res))
  }) %>% unlist()
  res_loc <- res_loc[worked]
  res[[as.character(multiplier)]] <- bind_rows(res_loc)
}
real <- bind_rows(res, .id = "multiplier") %>%
  clean_results()
```

```{r, eval = FALSE}
sce <- create_bifurcating_three_conditions_simu(multiplier = 1, nSim = 400)
set.seed(2019)
res_loc <- mclapply(seq_len(n_boot_null), function(b, frac) {
  return(anayze_multiple_conditions(sub_sample(sce, frac = frac)))
}, frac = frac_null, mc.cores = nCores)
worked <- lapply(res_loc, function(res){
  return(!"SingleCellExperiment" %in% class(res))
}) %>% unlist()
res_loc <- res_loc[worked]
null <- bind_rows(res_loc) %>%
  mutate(multiplier = 1) %>%
  clean_results()
```

```{r, eval = FALSE}
complex <- bind_rows("Real" = real, "Null" = null, .id = "Type") %>%
  arrange(multiplier)
```

## Trajectory with five lineages and two conditions

```{r, eval = FALSE}
ratio <- c(.1, .2, .3, .5, 2, 3, 5, 10)
res <- list()
for (multiplier in ratio) {
 sce <- create_5_lineages_simu(multiplier = multiplier)
 if (multiplier == .5) complex_sce <- sce
  set.seed(2019)
  res_loc <- mclapply(seq_len(n_boot_real), function(b, frac) {
    return(anayze_multiple_conditions(sub_sample(sce, frac = frac)))
  }, frac = frac_real, mc.cores = nCores)
  worked <- lapply(res_loc, function(res){
    return(!"SingleCellExperiment" %in% class(res))
  }) %>% unlist()
  res_loc <- res_loc[worked]
  res[[as.character(multiplier)]] <- bind_rows(res_loc)
}
real <- bind_rows(res, .id = "multiplier") %>%
  clean_results()
```

```{r, eval = FALSE}
sce <- create_5_lineages_simu(multiplier = 1, nSim = 400)
set.seed(2019)
res_loc <- mclapply(seq_len(n_boot_null), function(b, frac) {
  return(anayze_multiple_conditions(sub_sample(sce, frac = frac)))
}, frac = frac_null, mc.cores = nCores)
worked <- lapply(res_loc, function(res){
  return(!"SingleCellExperiment" %in% class(res))
}) %>% unlist()
res_loc <- res_loc[worked]
null <- bind_rows(res_loc) %>%
  mutate(multiplier = 1) %>%
  clean_results()
```

```{r, eval = FALSE}
five_lins <- bind_rows("Real" = real, "Null" = null, .id = "Type") %>%
  arrange(multiplier)
```

## Wrong step 1 outcome

```{r, eval = FALSE}
ratio <- c(.9, .95)
res <- list()
for (multiplier in c(ratio, round(1 / ratio, 2)[2:1])) {
  sce <- create_bifurcating_simu(multiplier = multiplier)
  set.seed(2020)
  res_loc <- mclapply(seq_len(n_boot_real), function(b, frac) {
    return(anayze_all_per_cond(sub_sample(sce, frac = frac)))
  }, frac = frac_real, mc.cores = nCores)
  worked <- lapply(res_loc, class) %>% unlist()
  worked <- worked == "data.frame"
  res_loc <- res_loc[worked]
  res[[as.character(multiplier)]] <- bind_rows(res_loc)
}
real <- bind_rows(res, .id = "multiplier")
```

```{r}
sce <- create_consecutive_bifurcating_simu(multiplier = 1, nSim = 400)
set.seed(2019)
null <- mclapply(seq_len(n_boot_null), function(b, frac) {
    return(anayze_all_per_cond(sub_sample(sce, frac = frac)))
  }, frac = frac_null, mc.cores = nCores) %>%
bind_rows()
```

```{r, eval = FALSE}
step1_fail <- bind_rows("Real" = real, "Null" = null, .id = "Type") %>%
  arrange(multiplier) %>%
  mutate(X = X - 1, group = X - X %% 5) %>%
  select(-X) %>%
  mutate(multiplier = as.numeric(multiplier))
```

## Unstability

```{r, eval = FALSE}
add_noise_to_sce <- function(sce, noise_pst = .1, noise_wgt = .1) {
  pseudotime <- matrix(sce$sim_time, ncol = 2,
                       nrow = ncol(sce), byrow = FALSE)
  weights <- matrix(c(sce$from %in% c("sA", "sB", "sBmid", "sC"),
                      sce$from %in% c("sA", "sB", "sBmid", "sD")),
                    nrow = ncol(sce), byrow = FALSE)
  normWeights <- sweep(weights, 1,
                       FUN = "/",
                       STATS = apply(weights, 1, sum))
  pseudotime <- pseudotime * matrix(rnorm(2 * ncol(sce), mean = 1, sd = noise_pst), ncol = 2)
  normWeights <- apply(normWeights, 1, function(row) {
    switch <- rbinom(1, 1, noise_wgt)
    if (switch) {
      return(1 - row)
    } else {
      return(row)
    }
  })
  return(list("pseudotime" = pseudotime,
              "cellWeights" = t(normWeights)))
}
```

```{r, eval = FALSE}
set.seed(20876)
sce <-  create_bifurcating_simu(multiplier = .85)
i <- 1
res <- list()
for (noise_pst in c(0:40 / 10)) {
  for (noise_wgt in c(0:5 / 10)) {
    for (j in 1:100) {
      print(i)
      estimates <- add_noise_to_sce(sce, noise_pst, noise_wgt)
      res_noise <- progressionTest(pseudotime = estimates$pseudotime, 
                                  cellWeights = estimates$cellWeights,
                                  conditions = sce$condition,
                                  thresh = .01) %>%
        select(-lineage)
      res_noise$noise_pst <- noise_pst
      res_noise$noise_wgt <- noise_wgt
      res[[i]] <- res_noise
      i <- i + 1
    }
  }
}
unstability <- bind_rows(res)
```

# Exploring the simulations
## Low dimension representation

```{r, fig.height=5, fig.width=6}
data("fork_sce", package = "condimentsPaper")
plot_reduced_dim_together(fork_sce) +
  annotate("text", x = 0, y = -.4, label = "Start", size = 4) +
  annotate("text", x = .35, y = .05, label = "Lineage 1", size = 4) +
  annotate("text", x = -.4, y = .15, label = "Lineage 2", size = 4)
```

```{r, fig.height=5, fig.width=6}
data("tree_sce", package = "condimentsPaper")
plot_reduced_dim_together(tree_sce)  +
  annotate("text", x = -.15, y = .35, label = "Start", size = 4) +
  annotate("text", x = -.3, y = -.1, label = "Lineage 1", size = 4) +
  annotate("text", x = .1, y = -.35, label = "Lineage 2", size = 4) +
  annotate("text", x = .25, y = -.05, label = "Lineage 3", size = 4)
```

```{r, fig.height=5, fig.width=6}
data("complex_sce", package = "condimentsPaper")
plot_reduced_dim_together(complex_sce) +
  annotate("text", x = -.15, y = -.3, label = "Start", size = 4) +
  annotate("text", x = .3, y = .2, label = "Lineage 1", size = 4) +
  annotate("text", x = -.35, y = .2, label = "Lineage 2", size = 4)
```

```{r, fig.height=5, fig.width=6}
data("fivelin_sce", package = "condimentsPaper")
plot_reduced_dim_together(fivelin_sce) +
  annotate("text", x = 0, y = -.1, label = "Start", size = 4) +
  annotate("text", x = .3, y = .15, label = "Lineage 1", size = 4, angle = 30) +
  annotate("text", x = .2, y = .3, label = "Lineage 2", size = 4, angle = 30) +
  annotate("text", x = -.3, y = .05, label = "Lineage 3", size = 4) +
  annotate("text", x = -.15, y = .3, label = "Lineage 4", size = 4) +
  annotate("text", x = 0, y = -.35, label = "Lineage 5", size = 4)
```

## Network branching
### Two lineages

```{r}
set.seed(0989021)
backbone <- backbone_bifurcating()
model_common <-
  initialise_model(
    backbone = backbone,
    num_cells = 100,
    num_tfs = nrow(backbone$module_info),
    num_targets = 250,
    num_hks = 250,
    simulation_params = simulation_default(
      census_interval = 10, 
      ssa_algorithm = ssa_etl(tau = 300 / 3600),
      experiment_params = simulation_type_wild_type(num_simulations = 100)
    )
  ) %>%
  generate_tf_network()
set.seed(298)
net <- plot_backbone_modulenet_simplify(model_common) +
  guides(col = FALSE, edge_width = FALSE) +
  ggraph::scale_edge_width_continuous(range = c(.5, .5)) + 
  theme(plot.background = element_blank()) 
net <- ggdraw() +
  draw_plot(net, 0, 0, scale = 1.2)
net
```

### Three lineages

```{r}
set.seed(1082)
backbone <- backbone_consecutive_bifurcating()
model_common <-
  initialise_model(
    backbone = backbone,
    num_cells = 100,
    num_tfs = nrow(backbone$module_info),
    num_targets = 250,
    num_hks = 250,
    simulation_params = simulation_default(
      census_interval = 10, 
      ssa_algorithm = ssa_etl(tau = 300 / 3600),
      experiment_params = simulation_type_wild_type(num_simulations = 100)
    )
  ) %>%
  generate_tf_network()
set.seed(2)
net2 <- plot_backbone_modulenet_simplify(model_common) +
  guides(col = FALSE, edge_width = FALSE) +
  ggraph::scale_edge_width_continuous(range = c(.5, .5)) + 
  theme(plot.background = element_blank()) 
net2 <- ggdraw() +
  draw_plot(net2, 0, 0, scale = 1.2)
net2
```

### Five lineages

```{r}
set.seed(15)
bblego_list <- list(
  bblego_start("A", type = "simple", num_modules = 2),
  bblego_linear("A", to = "B", num_modules = 2),
  bblego_branching("B", c("C", "D", "E")),
  bblego_branching("C", c("F", "G")),
  bblego_branching("E", c("H", "I")),
  bblego_end("D", num_modules = 4), 
  bblego_end("F", num_modules = 4), bblego_end("G", num_modules = 4),
  bblego_end("H", num_modules = 6), bblego_end("I", num_modules = 6)
)
backbone <- bblego(.list = bblego_list)
model_common <-
  initialise_model(
    backbone = backbone,
    num_cells = 100,
    num_tfs = nrow(backbone$module_info),
    num_targets = 250,
    num_hks = 250,
    simulation_params = simulation_default(
      census_interval = 10, 
      ssa_algorithm = ssa_etl(tau = 300 / 3600),
      experiment_params = simulation_type_wild_type(num_simulations = 100)
    )
  ) %>%
  generate_tf_network()
set.seed(14)
net3 <- plot_backbone_modulenet_simplify(model_common) +
  guides(col = FALSE, edge_width = FALSE) +
  scale_edge_width_continuous(range = c(.5, .5)) + 
  theme(plot.background = element_blank()) 
net3 <- ggdraw() +
  draw_plot(net3, 0, 0, scale = 1.2)
net3
```



# Analyzing the simulations

```{r}
data("fork", package = "condimentsPaper")
data("tree", package = "condimentsPaper")
data("complex", package = "condimentsPaper")
data("five_lins", package = "condimentsPaper")
```

## Comparison of methods

```{r}
df <- lapply(
  list("fork" = fork, "tree" = tree, "complex" = complex, "5lin" = five_lins),
  all_metrics, cutoff = .05) %>%
  bind_rows(.id = "dataset") %>%
  mutate(value = round(value, 2)) %>%
  mutate(metric = factor(metric, levels = c("TNR", "PPV", "TPR", "NPV", "F1")),
         dataset = case_when(
           dataset == "fork" ~ "a)",
           dataset == "tree" ~ "b)",
           dataset == "complex" ~ "c)",
           dataset == "5lin" ~ "d)"
         ),
         dataset = factor(dataset, levels = c("a)", "b)", "c)", "d)"))) %>%
  arrange(dataset, metric) %>%
  mutate(test_type = factor(test_type, 
                            levels = c("condiments_sling_prog", "condiments_mon_prog",
                                       "condiments_sling_diff", "DAseq", "milo")),
         label = if_else(is.na(value), "NA", as.character(value)))
table_plot <- function(df) {
  p <- ggplot(df, aes(x = test_type, y = metric, label = label,
               fill = as.numeric(label))) +
  geom_tile() +
  geom_text(na.rm = FALSE) +
  scale_fill_viridis_c() +
  scale_x_discrete(labels = c("progressionTest\n(Slingshot)", "progressionTest\n(Monocle)",
                              "differentationTest\n(Slingshot)", "DAseq", "Milo")) +
  labs(x = "", y = "", fill = "(Value - 1)\n") +
  guides(fill = "none") +
  facet_grid(dataset ~ ., switch = "y") +
  theme(axis.text.x = element_text(size = 14, angle = -20, vjust = .2)) +
  theme(legend.position =  "right",
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.background = element_blank(),
        panel.spacing.y = unit(7, "pt"),
        strip.text.y.left = element_text(angle = 0))
  return(p)
}
table_plot(df)
```


## Wrong step 1 outcome

```{r}
data("step1_fail", package = "condimentsPaper")
cutoffs <- c((2:9) * .001, (1:9) * .01, .1 * 1:10)
df <- step1_fail %>% 
  mutate(test_type = paste0(method, "_", test_type)) %>% 
  select(test_type, multiplier, p.value, Type) %>%
  dplyr::rename("adjusted_p_value" = "p.value") %>%
  lapply(cutoffs, all_metrics, df = .)
names(df) <- cutoffs
df <- bind_rows(df, .id = "cutoff") %>%
  mutate(cutoff = as.numeric(cutoff)) %>%
  select(cutoff, metric, value, test_type) %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  filter(str_detect(test_type, "sling")) %>%
  mutate(step1 = if_else(str_detect(test_type, "normal"),
                         "Fail to reject the null\n(Common trajectory)",
                         "Reject the null\n(Separate trajectory)"),
         test_type = word(test_type, 3, sep = "_"),
         test_type = if_else(test_type == "diff", "differentiationTest", "progressionTest"))
p <- ggplot(df, aes(x = 1 - PPV, y = TPR, col = step1)) +
  geom_path() +
  geom_point(size = 2) +
  facet_wrap(~test_type, scales = "free_y") + 
  scale_color_brewer(palette = "Set1") +
  labs(col = "Step 1 results:") +
  theme_classic() +
  theme(legend.position = "bottom")
p
```

## Unstability

```{r}
data("unstability", package = "condimentsPaper")
df <- unstability %>%
  group_by(noise_pst, noise_wgt) %>%
  summarise(p.value = median(p.value))
p <- ggplot(df, aes(x = noise_pst, y = p.value)) +
  geom_point(aes(col = as.character(noise_wgt))) +
  labs(x = "Noise on Pseudotime", y = "p value", 
       col = "Noise on\nbranch assignement")
legend <- get_legend(p + theme(legend.position = "bottom"))
p <- p + guides(col = FALSE, fill = FALSE)
plot_grid(plot_grid(p, p + scale_y_log10(), ncol = 2, scale = .95),
          legend, rel_heights = c(10, 1), nrow = 2)
```

## Special look at three condition situation

```{r}
ggplot(complex, aes(x = as.numeric(factor(multiplier, levels = unique(multiplier))),
                    y = adjusted_p_value)) +
  geom_boxplot(aes(group = interaction(test_type, multiplier),
                   fill = test_type)) +
  geom_hline(yintercept = .05) +
  scale_x_continuous(labels = unique(complex$multiplier),
                     breaks = 1:5) +
  scale_color_brewer(palette = "Pastel2") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(x = "Effect Size: Multiplier on the master regulator",
       y = "adjusted p-value", col = "Test", fill = "Test") +
  theme(legend.position = "bottom")
```

# References 
