---
title: 'Analysis of the TGFB datataset'
author: "Hector Roux de Bézieux, Koen Van den Berge and Kelly Street"
bibliography: ref.bib
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Analysis of the TGFB dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

We reproduce a lot of code from the [Bioc 2020 trajectory workshop](https://kstreet13.github.io/bioc2020trajectories/articles/workshopTrajectories.html).

# Dataset

The dataset we will be working with concerns a single-cell RNA-sequencing dataset consisting of two different experiments, which correspond to two treatments. McFaline-Figueroa et al. [@McFaline-Figueroa2019] studied the epithelial-to-mesenchymal transition (EMT), where cells spatially migrate from the epithelium to the mesenchyme during development.

```{r,include = FALSE}
libs <- c("dplyr", "tradeSeq", "SingleCellExperiment", "slingshot",
          "Seurat", "ggplot2", "condiments", "pheatmap", "fgsea", "msigdbr",
          "openxlsx")
suppressMessages(
  suppressWarnings(sapply(libs, require, character.only = TRUE))
)
rm(libs)
knitr::opts_chunk$set(fig.width=7)
theme_set(theme_classic())
```

# Load data

The data is imported using a function from the package. We then normalized using _Seurat_[@Stuart2019] and compute reduced dimension coordinates with UMAP [@Becht2019, @McInnes2018].

```{r, eval = FALSE}
tgfb <- condimentsPaper::import_TGFB()
library(Seurat)
########################
### Split by condition and convert to Seurat
########################
assays(tgfb)$logcounts <- log1p(assays(tgfb)$counts)
tgfbMock <- tgfb[ ,colData(tgfb)$pheno$treatment_id=='Mock']
tgfbTGFB <- tgfb[ ,colData(tgfb)$pheno$treatment_id=='TGFB']
soMock <- as.Seurat(tgfbMock)
soTGFB <- as.Seurat(tgfbTGFB)

########################
### Normalize
########################
soMock <- SCTransform(soMock, verbose = FALSE)
soTGFB <- SCTransform(soTGFB, verbose = FALSE)

########################
### Integrate
########################
dtlist <- list(Mock = soMock, TGFB = soTGFB)
intfts <- SelectIntegrationFeatures(object.list = dtlist, nfeatures = nrow(tgfb)) # maxes out at 4080 (why?)
dtlist <- PrepSCTIntegration(object.list = dtlist,
                             anchor.features = intfts)
anchors <- FindIntegrationAnchors(object.list = dtlist, normalization.method = "SCT",
                                  anchor.features = intfts)
integrated <- IntegrateData(anchorset = anchors, normalization.method = "SCT")
integrated <- RunPCA(integrated)
integrated <- RunUMAP(integrated, dims = 1:50)

## convert back to singleCellExperiment
tgfb <- as.SingleCellExperiment(integrated, assay = "RNA")
```

```{r}
data("tgfb", package = "condimentsPaper")
```

```{r compute scores, eval = TRUE, fig.width=7}
df <- bind_cols(
  as.data.frame(reducedDims(tgfb)$UMAP),
  as.data.frame(colData(tgfb))
  ) %>%
  sample_frac(1)
p1 <- ggplot(df, aes(x = UMAP_1, y = UMAP_2, col = pheno.treatment_id)) +
  geom_point(size = .7) +
  scale_color_brewer(palette = "Accent") +
  labs(col = "Treatment")
p1
```

```{r, fig.width=7}
p2 <- ggplot(df, aes(x = UMAP_1, y = UMAP_2, col = pheno.spatial_id)) +
  geom_point(size = .7) +
  scale_color_brewer(palette = "Dark2") +
  labs(col = "Spatial ID")
p2
```

```{r, eval = TRUE}
scores <- condiments::imbalance_score(
  Object = df %>% select(UMAP_1, UMAP_2) %>% as.matrix(), 
  conditions = df$pheno.treatment_id,
  k = 20, smooth = 40)
df$scores <- scores$scaled_scores
p3 <- ggplot(df, aes(x = UMAP_1, y = UMAP_2, col = scores)) +
  geom_point(size = .7) +
  scale_color_viridis_c(option = "C") +
  labs(col = "Scores")
p3
```

# Trajectory Inference and Differential Topology

To estimate the trajectory, we use _slingshot_ [@Street2018a].

```{r slingshot}
library(slingshot)
tgfb <- slingshot(tgfb, reducedDim = 'UMAP',
                  clusterLabels = colData(tgfb)$pheno$spatial_id,
                  start.clus = 'inner', approx_points = 100)
```

```{r}
set.seed(821)
topologyTest(SlingshotDataSet(tgfb), tgfb$pheno$treatment_id, rep = 100,
             methods = "KS_mean", threshs = .01)
```

```{r plotSlingshot}
df <- bind_cols(
  as.data.frame(reducedDims(tgfb)$UMAP),
  as.data.frame(colData(tgfb)[,-3])
  ) %>%
  sample_frac(1)
curve <- slingCurves(tgfb)[[1]]
p4 <- ggplot(df, aes(x = UMAP_1, y = UMAP_2, col = slingPseudotime_1)) +
  geom_point(size = .7) +
  scale_color_viridis_c() +
  labs(col = "Pseudotime") +
  geom_path(data = curve$s[curve$ord, ] %>% as.data.frame(),
            col = "black", size = 1.5)
p4
```

# Differential progression

```{r plotDensities}
p5 <- ggplot(df, aes(x = slingPseudotime_1)) +
  geom_density(alpha = .8, aes(fill = pheno.treatment_id), col = "transparent") +
  geom_density(aes(col = pheno.treatment_id), fill = "transparent",
               guide = FALSE, size = 1.5) +
  labs(x = "Pseudotime", fill = "Treatment") +
  guides(col = FALSE, fill = guide_legend(
    override.aes = list(size = 1.5, col = c("#7FC97F", "#BEAED4"))
  )) +
  scale_fill_brewer(palette = "Accent") +
  scale_color_brewer(palette = "Accent")
p5
```

```{r}
progressionTest(SlingshotDataSet(tgfb), conditions = tgfb$pheno$treatment_id)
```

# Differential expression

We use _tradeSeq_ [@VandenBerge2020].

## Select number of knots

```{r evaluateK, eval=FALSE}
library(tradeSeq)
set.seed(3)
icMat <- evaluateK(counts = as.matrix(assays(tgfb)$counts),
                   pseudotime = colData(tgfb)$slingshot$pseudotime,
                   cellWeights = colData(tgfb)$slingshot$cellWeights.V1,
                   conditions = factor(colData(tgfb)$pheno$treatment_id),
                   nGenes = 300,
                   k = 3:7)
```

## Fit GAM

```{r, eval=FALSE}
set.seed(3)
tgfb2 <- fitGAM(counts = tgfb, nknots = 5,
               conditions = factor(colData(tgfb)$pheno$treatment_id),
               )
mean(rowData(tgfb)$tradeSeq$converged)
```

## Differential expression between conditions

```{r, eval=TRUE}
condRes <- conditionTest(tgfb, l2fc = log2(2))
condRes$padj <- p.adjust(condRes$pvalue, "fdr")
mean(condRes$padj <= 0.05, na.rm = TRUE)
sum(condRes$padj <= 0.05, na.rm = TRUE)
conditionGenes <- rownames(condRes)[condRes$padj <= 0.05]
conditionGenes <- conditionGenes[!is.na(conditionGenes)]
```

## Visualize most and least significant gene

```{r, eval=TRUE}
library(RColorBrewer)
scales <- brewer.pal(3, "Accent")[1:2]

# plot genes
oo <- order(condRes$waldStat, decreasing = TRUE)

# most significant gene
p6 <- plotSmoothers(tgfb, assays(tgfb)$counts,
              gene = rownames(assays(tgfb)$counts)[oo[1]],
              alpha = 1, border = TRUE, curvesCols = scales) +
  scale_color_manual(values = scales) +
  ggtitle(rownames(assays(tgfb)$counts)[oo[1]])

# Second most significant gene
p7 <- plotSmoothers(tgfb, assays(tgfb)$counts,
              gene = rownames(assays(tgfb)$counts)[oo[2]],
              alpha = 1, border = TRUE, curvesCols = scales) +
  scale_color_manual(values = scales) +
  ggtitle(rownames(assays(tgfb)$counts)[oo[2]])

# least significant gene
p8 <- plotSmoothers(tgfb, assays(tgfb)$counts,
              gene = rownames(assays(tgfb)$counts)[oo[nrow(tgfb)]],
              alpha = 1, border = TRUE, curvesCols = scales) +
  scale_color_manual(values = scales) +
  ggtitle(rownames(assays(tgfb)$counts)[oo[nrow(tgfb)]])
p6
p7
p8
```

## Heatmaps of genes DE between conditions

Below we show heatmaps of the genes DE between conditions. The DE genes in the heatmaps are ordered according to a hierarchical clustering on the TGF-Beta condition.

```{r, eval=TRUE}
library(cowplot)
library(scales)
### based on mean smoother
yhatSmooth <- predictSmooth(tgfb, gene = conditionGenes, nPoints = 50, tidy = FALSE) %>%
  log1p()
yhatSmoothScaled <- t(apply(yhatSmooth,1, scales::rescale))
heatSmooth_TGF <- pheatmap(yhatSmoothScaled[, 51:100],
  cluster_cols = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "TGF-Beta", legend = FALSE,
  silent = TRUE
)

matchingHeatmap_mock <- pheatmap(yhatSmoothScaled[heatSmooth_TGF$tree_row$order, 1:50],
  cluster_cols = FALSE, cluster_rows = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "Mock",
  legend = FALSE, silent = TRUE 
)

p9 <- plot_grid(heatSmooth_TGF[[4]], matchingHeatmap_mock[[4]], ncol = 2)
p9
```

## Gene set enrichment analysis

This is done using the _fgsea_ package [@Korotkevich2016].
```{r}
geneSets <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "BP") %>%
  mutate(gene_symbol = toupper(gene_symbol)) %>%
  filter(gene_symbol %in% names(tgfb)) 
m_list <- geneSets %>% split(x = .$gene_symbol, f = .$gs_name)
statsCond <- condRes$waldStat
names(statsCond) <- rownames(condRes)
eaRes <- fgsea(pathways = m_list, stats = statsCond, nperm = 5e4, minSize = 10)
eaRes <- eaRes %>% arrange(pval)
knitr::kable(head(eaRes[, 1:3], n = 20))
```

# Session info 

```{r}
sessionInfo()
```

# References
